Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2013-10-19 ieee_paper.tex
shrinking text

Each vector in the problem is of dimension $n_r$ and each entry
takes $b_x$ bytes. There are $n_v$ {\tt x} vectors and $n_v n_m$ {\tt
  y} vectors, which lead to the size of the {\tt x} and {\tt y} vectors:
$$s_x = n_v b_x n_r$$ $$s_y = n_v n_m b_x n_r$$

@@

In this work, we use an Intel Xeon Phi 5110P coprocessor. This card
has 8 memory controllers that can each execute 5 billion
transactions per second. Each has two 32-bit channels, achieving a total
bandwidth of 320~GB/s aggregated across all the memory
controllers. There are 60 cores clocked at 1.05~GHz. Their memory
interfaces are 32-bit wide with two channels and the total bandwidth
is 8.4~GB/s per core. Thus, the cores should be able to consume 504~GB/s
at most. However, the bandwidth between the cores and the memory
controllers is limited by the ring network that connects the cores and
the memory controller. Its precise bandwidth is unknown but it is
believed to be between 200~GB/s and 250~GB/s.

@@

Most of the performance of the architecture comes from the vector
processing unit. Each core has 32 512-bit SIMD registers that can be
used for double or single precision, that is, either as a vector of 8
64-bit values or as a vector of 16 32-bit values, respectively. The
vector processing unit can perform many basic instructions, such as
addition or division, and mathematical operations, such as {\tt sin} and
{\tt sqrt}, making it possible to reach 16 single precision (or 8 double precision)
operations per cycle.  
The unit also supports Fused Multiply-Add (FMA)
operations, which are typically counted as two operations for
when benchmarking. Therefore, the peak performance of the 5110P
card is 1 Tflop/s in double precision and 2 Tflop/s in single
precision. If FMA cannot be used, only half of these rates can be
achieved.

@@

As explained in Section~\ref{sec:model}, we expect to achieve a performance of
about 210~Gflop/s in single precision in the best case. We will focus
on the single precision case with four vectors and four matrices. 
Similar techniques apply for other matrix/vector combinations. This target
performance represents 10\% of the peak performance of the
architecture which is impossible to reach without an efficient vectorization of
the kernel. Indeed, to reach such a performance, a Fused Multiply-Add
instruction on fully loaded registers must be executed at most every
10 cycles.

@@

Loading the entries from the {\tt x} vector is a little bit more
involved. First, the columns pointers of four different nonzeros are
loaded into a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation\footnote{Notice that there are two calls to {\tt unpack} for the
  LSB and MSB. Both are mandatory even if one is known to be
  nilpotent according the documentation of the hardware}. (The column
pointers are named $j,k,l,m$ in Figure~\ref{code:mat_mul}.) They
are replicated so that each value fills an single vector lane 
using a swizzle operation. Then each value is multiplied by 4 
(because there are 4 vectors) 
and the offset vector $(3,2,1,0)$ is added to each lane to obtain the
correct index of the elements of the {\tt x} vector within each lane
using a Fused Multiply-Add operation. Then a {\tt gather} operation is
performed to bring the 16 entries of the {\tt x} vector into the SIMD
register. At this point we have a SIMD register where each lane is the
four vector entries for a nonzero element. Using a broadcast
permutation, one can replicate an entry of the four vectors across
each lane. (The broadcast of $x_m^1,x_m^2,x_m^3,x_m^4$ is shown in
Figure~\ref{code:mat_mul}.)

@@

\section{Experimental Validation}
\label{sec:expe}
In this section, we describe a series of numerical experiments to
confirm the performance predictions of Section~\ref{sec:model}. To
better understand our results, we first perform some bandwidth
measurements, followed by the actual SpMV computations. All the codes
are compiled with the Intel C Compiler version 13 and
-O3 optimization. The codes use the IMCI instructions exposed by
the compiler through intrinsics.

@@

\begin{figure}[bth]
  \centering
  \includegraphics[width=.9\linewidth]{figures/bandwidth_read_write.pdf}

>>>>

Each vector in the problem is of dimension $n_r$ and each entry
takes $b_x$ bytes. There are $n_v$ {\tt x} vectors and $n_v n_m$ {\tt
  y} vectors, which lead to the size of the {\tt x} and {\tt y} vectors:
$$s_x = n_v b_x n_r\mbox{, ~}s_y = n_v n_m b_x n_r$$

@@

In this work, we use an Intel Xeon Phi 5110P coprocessor. This card
has 8 memory controllers that can each execute 5 billion
transactions per second. Each has two 32-bit channels, achieving a total
bandwidth of 320~GB/s aggregated across all the memory
controllers. There are 60 cores clocked at 1.05~GHz. Their memory
interfaces are 32-bit wide with two channels and the total bandwidth
is 8.4~GB/s per core. Thus, the cores should be able to consume 504~GB/s
at most. However, the bandwidth between the cores and the memory
controllers is limited by the ring network that connects the cores and
the memory controller. Its precise bandwidth 
believed to be between 200~GB/s and 250~GB/s.

@@

Most of the performance of the architecture comes from the vector
processing unit. Each core has 32 512-bit SIMD registers that can be
used for double or single precision, that is, either as a vector of 8
64-bit values or as a vector of 16 32-bit values, respectively. The
vector processing unit can perform basic instructions, such as
addition or division, and mathematical operations, such as {\tt sin} and
{\tt sqrt}, making it possible to reach 16 single precision (or 8 double precision)
operations per cycle.  
The unit also supports Fused Multiply-Add (FMA)
operations, which are typically counted as two operations for
when benchmarking. Therefore, the peak performance of the 5110P
card is 1 Tflop/s in double precision and 2 Tflop/s in single
precision. If FMA cannot be used, only half of these rates can be
achieved.

@@

As explained in Section~\ref{sec:model}, we expect to achieve a performance of
about 210~Gflop/s in single precision in the best case. We will focus
on the single precision case with four vectors and four matrices. 
Similar techniques apply for other combinations. This target
performance represents 10\% of the peak performance of the
architecture which cannot be reached without an efficient vectorization. Indeed, to reach such a performance, a Fused Multiply-Add
instruction on fully loaded registers must be executed at most every
10 cycles.

@@

Loading the entries from the {\tt x} vector is a little bit more
involved. First, the columns pointers of four different nonzeros are
loaded into a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation\footnote{Notice that there are two calls to {\tt unpack} for the
  LSB and MSB. Both are mandatory even if one is known to be
  nilpotent according the documentation of the hardware}. (The column
pointers are named $j,k,l,m$ in Figure~\ref{code:mat_mul}.) They
are replicated so that each value fills an single vector lane 
using a swizzle operation. Then each value is multiplied by 4 
(because $n_v=4$) 
and the offset vector $(3,2,1,0)$ is added to each lane to obtain the
correct index of the elements of the {\tt x} vector within each lane
using a Fused Multiply-Add operation. Then a {\tt gather} operation is
performed to bring the 16 entries of the {\tt x} vector into the SIMD
register. At this point we have a register where each lane is the
four vector entries for a nonzero element. Using a broadcast
permutation, one can replicate an entry of the four vectors across
each lane. (The broadcast of $x_m^1,x_m^2,x_m^3,x_m^4$ is shown in
Figure~\ref{code:mat_mul}.)

@@

\section{Experimental Validation}
\label{sec:expe}

\begin{figure}[tbh]
  \centering
  \includegraphics[width=.9\linewidth]{figures/bandwidth_read_write.pdf}

In this section, we describe a series of numerical experiments to
confirm the performance predictions of Section~\ref{sec:model}. To
better understand our results, we first perform some bandwidth
measurements, followed by the actual SpMV computations. All the codes
are compiled with the Intel C Compiler version 13 and
-O3 optimization. The codes use the IMCI instructions exposed by
the compiler through intrinsics.

@@
